### 2016-04-14 Manuscript/SW_early_life_telomeres.Rmd
More post-Dave edits

To assess our resolution for detecting telomere shortening, we tested, for each age class, whether $\Delta$RTL was significantly lower than zero, using one-sample t-tests. We then tested how $\Delta$RTL was related to the time interval between sampling events, using linear models. A significantly negative slope indicates that telomere shortening occurs linearly over time, while the time point at which $\Delta$RTL is significantly lower than zero can be considered the time period over which we have the resolution to detect reductions in telomere length.

@@

Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`). Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C). Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`), and did not differ between birds born in minor compared to main breeding seasons (t = `r round2(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Season,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`; Fig. 2A, 2C).

@@

#####Early-life telomere shortening
Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig. 4A). $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`). In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase. In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`). However, in fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`; Fig. 4B, Fig. S1). In both fledglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 4B). Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life. We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings (N = `r nrow(subset(Loss,Agemonths == 1))`).

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`), or sex (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Sex+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[5,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Sex+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[5,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).

>>>>

@@

Early-life RTL varied significantly among cohorts (Likelihood ratio test: F = `r round2(anova(lme(RTL~factor(FieldPeriodID),data=juv,random=~1|BirdID))[2,3])`; P `r round3(anova(lme(RTL~Agemonths,data=juv,random=~1|BirdID))[2,4])`). Cohort-level variation in juvenile RTL was positively related to adult population density (LMM controlling for age and cohort: t = `r round2(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Density,data=juv,random=~1|FieldPeriodID))$tTable[3,5])`; Figs 2B,2C). Variation in cohort-level RTL was not related temporal variation in  food availability (t = `r round2(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,4])`; P `r round3(summary(lme(RTL~Agemonths+Insect,data=juv,random=~1|FieldPeriodID,na.action=na.exclude))$tTable[3,5])`).

@@

#####Early-life telomere shortening
Longitudinal data showed that early-life RTL was not related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig. 4A). $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`). In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`; Fig. 4B), suggesting that telomere shortening was highest shortly after the nestling phase.

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[4,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cohort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+RTL+Agemonths+(1|TerritoryID)+(1|FieldPeriodID),data=juv13,family = 'binomial'))$coefficients[2,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data = subset(Loss,FieldPeriodID<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~Loss+Agemonths,data=subset(Loss,LayYear<2013),family = 'binomial'))$coefficients[2,4])`).

### 2016-04-14 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
More post-Dave edits

temp <- juv
x <- ifelse(temp$Season == 'Minor',0.5,0)
temp$LayYear2 <- temp$LayYear+x
temp$LayYear <- factor(temp$LayYear)

Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL,fill=Season))+
  ylab('RTL') + 
  xlab('Cohort') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12,angle = 90), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top')+
  geom_jitter(aes(colour = Season),position=position_jitterdodge(dodge.width = 0.8))+
  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
  scale_colour_manual(values = c('grey','darkorange'))+
  scale_fill_manual(values = c('grey','darkorange'))+
  annotate('text',x = 1,y = 2.1,label = 'A')

temp <- temp[!(duplicated(temp$LayYear2)),]

Fig2B <- ggplot(temp,aes(x = LayYear2,y = Density))+
  ylab('Adult population size') + 
  xlab('Cohort') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'none')+
  geom_line(stat = 'identity')+
  annotate('text',x = 1998,y = 360,label = 'B')

temp <- juv
temp[,'Age (months)'] <- factor(temp$Agemonths)

Fig2C <- ggplot(temp,
       aes(x = Density,
           y = RTL,
           colour=Season,
           fill = Season)) +
  ylab('RTL') + 
  xlab('Adult population size') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'none') +
  geom_point()+
  stat_smooth(method = 'lm')+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
  annotate('text',x = 250,y = 2.1,label = 'C')

pdf('Figures/Figure 2.pdf',
    width = 8,
    height = 8,
    paper = 'a4r')
multiplot(Fig2A,Fig2B,Fig2C,layout=matrix(c(1,1,2,3),nrow = 2,byrow=T))
dev.off()

@@

Fig3C <- ggplot(temp,
       aes(x = `Age (months)`,
           y = RTL,
           fill=Sex)) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_jitter(aes(colour = Sex),position=position_jitterdodge(dodge.width = 0.8))+
  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
  scale_colour_manual(values = c('grey','darkorange'))+
  scale_fill_manual(values = c('grey','darkorange'))+
  annotate('text',x = 0.55,y = 2.1,label = 'C')

@@

temp <- Loss
temp[,'Age (months)'] <- factor(temp$Agemonths)
temp$LayYear <- factor(temp$LayYear)

Fig4A <- ggplot(temp,
       aes(x = RTL,
           y = RTL1,
           col=Ageclass)) +
  ylab('Adult RTL') + 
  xlab('Juvenile RTL') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_point()+
  stat_smooth(method = 'lm')+
  annotate('text',x = 0,y = 2.1,label = 'A')

temp2 <- temp
temp2$TimeDiff <- temp2$TimeDiff/365

Fig4B <- ggplot(temp2,
       aes(x = Ageclass,
           y = Loss)) +
  ylab(expression(Delta*'RTL')) + 
  xlab('Time (years)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(col = 'grey')+
  annotate('text',x = 0.5,y = 1.5,label = 'B')+
  geom_abline(slope = 0,intercept = 0,lty = 2)

@@

#Sort out row/col names
nams <- rownames(table2)
nams[nams == 'z.NonHelper'] <- 'Non helpers'
nams[nams == 'c.Sex'] <- 'Sex (males)'
nams[nams == 'z.TQ'] <- 'TQ'
nams[nams == 'c.Helper'] <- 'Helpers'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Agemonths'] <- 'Age'
nams[nams == 'z.Tarsus'] <- 'Tarsus'

@@

Fig5A <- forestplot(table2)+
  annotate('text',x = 0.75,y = 0.45,label = 'A')

boxplot(Loss~Helper*Ageclass,data=Loss)

temp <- Loss

Fig5B <- ggplot(temp,
       aes(col = factor(Helper),
           y = Loss,
           x=Ageclass)) +
  ylab(expression(Delta*'RTL')) + 
  xlab('Tarsus length (mm)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_boxplot(notch=T)+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))

pdf('Figures/Figure 5.pdf',
    width = 4,
    height = 8,
    paper = 'a4')
multiplot(Fig5A,Fig5B)
dev.off()

ggplot(juvseason9,aes(x = RTL,y = Lifespan,col = Season))+
  geom_point()+
  stat_smooth(method = 'lm')

>>>>

temp <- juv
temp$LayYear <- factor(temp$LayYear)

Fig2A <- ggplot(temp,aes(x = LayYear,y = RTL))+
  ylab('RTL') + 
  xlab('Cohort') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12,angle = 90), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top')+
  geom_jitter(col = 'grey')+
  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
  scale_colour_manual(values = c('grey','darkorange'))+
  scale_fill_manual(values = c('grey','darkorange'))+
  annotate('text',x = 1,y = 2.1,label = 'A')

temp$LayYear <- as.numeric(paste(temp$LayYear))

Fig2B <- ggplot(temp,aes(x = LayYear,y = Density))+
  ylab('Adult population size') + 
  xlab('Cohort') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'none')+
  geom_line(stat = 'identity')+
  annotate('text',x = 1998,y = 360,label = 'B')

pdf('Figures/Figure 2.pdf',
    width = 8,
    height = 6,
    paper = 'a4r')
multiplot(Fig2A,Fig2B,layout=matrix(c(1,2),nrow = 2,byrow=T))
dev.off()

@@

Fig3C <- ggplot(temp,
       aes(col = `Age (months)`,
           fill =  `Age (months)`,
           y = RTL,
           x=TQ)) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_point()+
  stat_smooth(method = 'lm')+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))+
  annotate('text',x = 3,y = 2.1,label = 'C')

@@

temp <- Loss
temp[,'Age (months)'] <- factor(temp$Agemonths)

Fig4A <- ggplot(temp,
       aes(x = RTL,
           y = RTL1,
           col=`Age (months)`)) +
  ylab('Adult RTL') + 
  xlab('Juvenile RTL') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_point()+
  scale_colour_manual(values = c('grey','darkorange','darkgreen'))+
  scale_fill_manual(values = c('grey','darkorange','darkgreen'))
  annotate('text',x = 0,y = 2.1,label = 'A')

Fig4B <- ggplot(temp,
       aes(x = `Age (months)`,
           y = Loss)) +
  ylab(expression(Delta*'RTL')) + 
  xlab('Age (months)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(col = 'grey')+
  annotate('text',x = 0.5,y = 1.5,label = 'B')+
  geom_abline(slope = 0,intercept = 0,lty = 2)

@@

@@

Fig5A <- forestplot(table2)+
  annotate('text',x = 0.75,y = 0.6,label = 'A')

temp <- Loss
temp[,'Helpers'] <- ifelse(temp$Helper>0,'Present', 'Absent')
temp[,'Non helpers'] <- ifelse(temp$NonHelper>0,'Present', 'Absent')
temp[,'Age (months)'] <- factor(temp$Agemonths)

Fig5B <- ggplot(temp,
       aes(x = `Age (months)`,
           y = Loss,
           fill=Helpers)) +
  theme_classic() +
  ylab(expression(Delta*'RTL'))+
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = 'top') +
  geom_jitter(aes(colour = Helpers),position=position_jitterdodge(dodge.width = 0.8))+
  geom_boxplot(outlier.colour = NA,notch=F,alpha=0.2)+
  scale_colour_manual(values = c('grey','darkorange'))+
  scale_fill_manual(values = c('grey','darkorange'))+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  annotate('text',x = 0.55,y = 1.5,label = 'C')

pdf('Figures/Figure 5.pdf',
    width = 8,
    height = 8,
    paper = 'a4')
multiplot(Fig5A,Fig5B)
dev.off()
